What, if anything, is epsilon?

### 2014-03-01 epsilon/paper.tex
Submitted version of Epsilon paper

% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}

@@

@@

\begin{abstract}
what do I put here
\end{abstract}

@@

Most programmers find this subject too tedious and simply pick a
number that seems pretty small. Thus in practice, $\epsilon$ is an
application-specific choice. Of well-known ``constants,''
$\epsilon$ may be the least agreed-upon, with values seen in the wild
spanning {\em 300 orders of magnitude}. This paper explores the practical
values of $\epsilon$ in real software.

@@

Did you know that the SPEC benchmarks~\cite{spec2006} cost \$800? Like
they literally expect me to pay them money to download the source code
so that I could grep for {\tt const double epsilon} or test my
compiler out on that. Mmany are even based on open-source software like
Sphinx and POV-Ray. Ridiculous. I refuse. Values of epsilon for the
SPEC benchmarks do not appear in Figure~\ref{fig:spec}.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{chart-c-double}
\end{center}\vspace{-0.1in}
\caption{ Values for {\tt const double epsilon} in the C programming
  language. In this chart, the blue (lower) bars are the distinct values
  of epsilon seen. The vertically aligned red (upper) bar is its count.
  Epsilon values are plotted on a logarithmic scale, where the minimum
  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
  $\log(1.609)$ is $5$. 
  %
  Notes: One programmer used the value \tt{-1e10}, which is
  -10,000,000,000, problably meaning \tt{1e-10}. This value was excluded
  because it has no real logarithm.
}
\label{fig:cdouble}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
% \includegraphics[width=0.99 \linewidth]{chart-c-double}
\end{center}\vspace{-0.1in}
\caption{
  Values for {\tt const double epsilon} and {\tt constexpr
    double epsilon} in the C++ programming language. The {\tt
    constexpr} qualifier, a new feature of C++11, is used very rarely
  (less than 1\% of the time).
  Notes:
  Five times, the computed value is actually equal to 0.0,

  The value {\tt pow(10,-13)} is annotated in German
  ``Genauigkeitsziel bei der Nullstellensuche,'' or ``Accuracy goal in
  the search for zeros.''
}
\label{fig:cppdouble}
\end{figure}

>>>>

% \usepackage{eclbkbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{code}
\usepackage{url}
% \usepackage{amscd}
% \usepackage{xy}
\usepackage{graphicx}

@@

\newcommand\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}

@@

\begin{abstract}
We present a sample of the values of the programming constant {\tt
  epsilon} as found on the internet, for several different programming
languages and with a variety of visualizations.
\end{abstract}

@@

Most programmers find this subject too tedious and simply pick a
number that seems pretty small. Thus in practice, $\epsilon$ is an
application-specific choice. Of well-known ``constants,'' $\epsilon$
may be the least agreed-upon, with values seen in the wild spanning
more than {\em 300 orders of magnitude}. This paper explores the
practical values of $\epsilon$ in real software.

@@

\paragraph{Github.} Github is the hub that contains all gits, approximately
10~million of them, as of the beginning of 2014. The site has search
functionality, which ``allowed'' me to scrape one hundred pages of
results for queries like {\tt "const double epsilon ="} for various
languages, as long as I didn't do it too fast. I scraped the
programming languages C, \CC, C$\sharp$, JavaScript, and Objective~C.
Each language has its own idiosyncracies about how constants are
defined, so I used one (or several) appropriate to each language. For
example, in JavaScript, I looked for {\tt "var epsilon ="}. From these
HTML files I extracted all of the right-hand-side expressions,
manually excluded the ones that could not be evaluated (for example
because they depended on other symbols; see
Figure~\ref{fig:cdoubleuninterpretable} for some examples), and then
computed the actual values for the rest. The source code to do the
scraping, extract the expressions, and tally the results is available
online.\!\footnote{In the Subversion repository at:
  \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/epsilon/}}

\paragraph{SPEC benchmarks.} Did you know that the SPEC 
benchmarks~\cite{spec2006} cost \$800? Like they literally expect me
to pay them money to download the source code so that I could grep for
{\tt const double epsilon} or test my compiler out on that. Many are
even based on open-source software like Sphinx and POV-Ray.
Ridiculous. I refuse. Values of epsilon for the SPEC benchmarks do not
appear in Figure~\ref{fig:spec}.

@@

The results of the analysis appear in several figures which are
interspersed haphazardly with this text. Each figure presents the data
in a different way, since this diversity in presentation should
maximize the chance that one of the charts makes sense to you. The C
programming language has two numeric types that could reasonably be
used to represent $\epsilon$: {\tt float} and {\tt double}. The
results for {\tt double} appear in Figure~\ref{fig:cdouble} and for
{\tt float} in Figure~\ref{fig:cfloat}. \CC\ has those same two types,
but I decided arbitrarily to only look at {\tt double}, which is in
Figure~\ref{fig:cppdouble}. Programmers in C$\sharp$ are very
creative; their results are presented in Figure~\ref{fig:csharp}.
Objective~C proved unpopular for use of $\epsilon$, its sparse data
are in Figure~\ref{fig:objectivec}. Finally, the ineffable JavaScript
has its results in Figure~\ref{fig:javascript}.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{chart-c-double}
\end{center}\vspace{-0.1in}
\caption{ Values for {\tt const double epsilon} in the C programming
  language. In this chart, the blue (lower) bars are the distinct values
  of epsilon seen. The vertically aligned red (upper) bar is its count.
  Epsilon values are plotted on a logarithmic scale, where the minimum
  observed value $\log(-708)$ is $303 \times 10^{-308}$, and the largest
  $\log(1.609)$ is $5$. 
  %
  Notes: One programmer used the value {\tt -1e10}, which is
  -10,000,000,000, probably meaning {\tt 1e-10}. This value was excluded
  because it has no real logarithm.
}
\label{fig:cdouble}
\end{figure}

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{chart-cpp}
\end{center}\vspace{-0.1in}
\caption{
  Values for {\tt const double epsilon} and {\tt constexpr
    double epsilon} in the C++ programming language. The {\tt
    constexpr} qualifier, a new feature of C++11, is used very rarely
  (less than 1\% of the time).
  %
  Notes:
  Five times, the programmer used 0.0 for epsilon, which is possibly
  the only unjustifiable value. The value {\tt pow(10,-13)} is
  annotated in German ``Genauigkeitsziel bei der Nullstellensuche,''
  or ``Accuracy goal in the search for zeros.'' }
\label{fig:cppdouble}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{chart-csharp}
\end{center}\vspace{-0.1in}
\caption{ Values for {\tt const double epsilon} in the C$\sharp$
  programming language. In this chart, the left edge of greyscale
  blocks on the left-facing flag represent the cumulative value of the
  logarithm of all values of epsilon seen so far. The right edge of
  the block with the exact same greyscale value on the bottom-right
  flag is the cumulative count of all definitions of epsilon so far.
  It is easy to see that there are 1,000 total definitions, as
  expected.
  %
  Notes: One programmer defined epsilon as {\tt 1.0-18}, that is,
  -17, probably meaning {\tt 1.0e-18}. Four used literal {\tt 0} for
  epsilon. Another defined epsilon as {\tt 1 / 100000}, which uses
  integer division and results again in 0. Owing to the strength and
  creativity of C$\sharp$ developers, here we saw our largest
  value of epsilon so far, {\tt 700}, and the declaration with the most
  significant digits: {\tt
    0.0000000000000002220446049250313080847263336181640625}.
}
\label{fig:csharp}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.99 \linewidth]{chart-objectivec}
\end{center}\vspace{-0.1in}
\caption{Results for the language Objective~C for {\tt const double epsilon}.
Perhaps because Objective~C programmers do not use this idiom for defining
constants, or because it is only used to make iPhone games where epsilon is
not a concept of interest, there were only 8 distinct values of epsilon
observed. This allows us to present the data completely in the chart. The
$x$ axis is $\sin(\epsilon)$ and the $y$ axis is $\cos(f)$ where $f$ is
the total number of times that the given value occurred in the code.}
\label{fig:objectivec}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.70 \linewidth]{chart-js}
\end{center}\vspace{-0.1in}
\caption{Finally, JavaScript, the official language of the Internet.
  In this ``radar'' plot, the data are arranged around a circle, as
  tuples of $\langle \log(\epsilon), \epsilon, f \rangle$ where $f$ is
  the frequency of the value being observed. Numbers in boxes ascend
  along the $0\deg$ axis, displaying each of the multiples of 5
  between 0 and 100, inclusive. Boxes stack so that only part of the
  number is visible, but you know what's under there if you've looked
  at numbers before. A spider-web-like network of interlacing lines
  ascribe some additional meaning to some of the points on the clock
  face. The smallest nonzero value observed was $10^{-32}$.
%
  Notes: JavaScript programmers have the highest tolerance for error
  of all languages tested, with over 1,000 using epsilon of 0.5 or
  higher. One programmer used $\epsilon = 1024$, and another
  6,000,000! }
\label{fig:javascript}
\end{figure}
